---
title: "Supplementary material (S1)"
subtitle: "Modelling Relation and SEM application to SDMs"
author: 
      - Daniele Da Re (ELIC - UCLouvain; BE)
      - Enrico Tordoni (University of Tartu; EST)
      - Jonathan Lenoir (Université de Picardie; FRA)
date: "`r format(Sys.time(), '%d/%m/%y')`"
output:
  pdf_document:
    fig_crop: no
    toc: true
    toc_depth: 3
always_allow_html: yes
header-includes: 
  - \renewcommand{\and}{\\}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      dev = 'pdf', 
                      out.width = "100%")
```
```{r, eval=TRUE, echo=FALSE }
setwd("/home/ddare/working_files/ENM_MR/SEM/script/")
# webshot::install_phantomjs()
```

## 1. Project overview 
Conceptual and technical application of Rosen's Modelling relations approach to Structural Equation Modelling (SEM) and Species Distribution Models. We simulated two virtual species (a virtual tree species and a virtual herbaceous plant) using two climatic variables (BIO1 - mean annual temperature, and BIO12 - annual precipitation) and one functional trait (i.e., seed germination rate).  

Load auxiliary function for rescaling datasets.
```{r, eval=TRUE, message=FALSE}
rescale <- function(x, x.min = NULL, x.max = NULL, new.min = 0, new.max = 1) {
  if(is.null(x.min)) x.min = min(x)
  if(is.null(x.max)) x.max = max(x)
  new.min + (x - x.min) * ((new.max - new.min) / (x.max - x.min))
}
```

## 2. Virtual species
```{r, eval=TRUE, message=FALSE}
library(tidyverse)
library(ggplot2)
library(virtualspecies)
library(raster)
library(knitr)
```

### First virtual species

Create a virtual "tree" species. 

```{r, eval=TRUE, message=FALSE}
Worldclim <- raster::getData('worldclim', var='bio', res = 10) 
envData <- crop(Worldclim, extent(-12, 25, 36, 60))
b <- as(extent(envData), "SpatialPolygons")
rm(Worldclim)

my.parameters <- formatFunctions(bio1 = c(fun = 'dnorm', mean = 100, sd = 400),
                                 bio12 = c(fun = 'dnorm', mean = 2000, sd = 4000))

my.first.species <- generateSpFromFun(raster.stack = envData[[c("bio1", "bio12")]],
                                      parameters = my.parameters,
                                      plot = FALSE)

plot(my.first.species$suitab.raster, main = "Virtual tree \n climatic suitability")
```





### Second virtual species


#### Development of germination rate function depending on the virtual tree suitability
\hfill\break

Since we want to simulate a virtual herbaceous species which is shade-tolerant and it grows underneath the virtual tree species, we defined an exponential function to predict the seed germination probability of the herbaceous species as a function of the "shading" effect provided by the virtual tree species. In short, we assume that at a higher tree suitability corresponds a higher germination probability of the herbaceous species.

We then create a shading germination layer based on the virtual tree climatic suitability
```{r, eval=TRUE}
fake_tree_suitab <- seq(0,1, by = 0.01)  
a <- 1         
g.r <- 1.5    
shade.germ <- a*exp(g.r*fake_tree_suitab)
shade.germ <- rescale(shade.germ, 
                     x.min = min(shade.germ), 
                     x.max = max(shade.germ), 
                     new.min = 0, new.max = 1)
plot(fake_tree_suitab, 
     shade.germ, 
     xlab = "Virtual tree climatic suitability", 
     ylab = "Germination rate")
```

 and compute a germination probability raster.
```{r, eval=TRUE}
shade.germ.r <- a*exp(g.r*my.first.species$suitab.raster) 
shade.germ.r <- rescale(shade.germ.r, x.min = 1, x.max = 5, new.min = 0, new.max = 1)

plot(shade.germ.r, main = "Germination rate of the virtual herb species")
```





#### Virtual "herbaceous plant" species
\hfill\break
We thus create the virtual herbaceous species whose distribution is restricted by the germination probability layer ad described above.

Note that that the climatic niche of the virtual herbaceous species is rather wide (i.e., ubiquitous distribution), but the actual "habitat" suitability will be restricted by the germination rate. 

```{r, eval=TRUE, message=FALSE}
my.parameters <- formatFunctions(bio1 = c(fun = 'dnorm', mean = 12, sd = 500),
                                 bio12 = c(fun = 'dnorm', mean = 1200, sd = 2000))

my.second.species <- generateSpFromFun(raster.stack = envData[[c("bio1", "bio12")]],
                                       parameters = my.parameters,
                                       formula = "bio1 + bio12",
                                       plot = FALSE)
plot(my.second.species$suitab.raster, 
     main = "Virtual herbaceous plant \n climatic suitability")
```



```{r, eval=TRUE, message=FALSE}
par(oma=c(3,3,3,3))
plotResponse(my.first.species)
title(sub = "Virtual tree species")
```


```{r, eval=TRUE, message=FALSE}
par(oma=c(3,3,3,3))
plotResponse(my.second.species)
title(sub = "Virtual herbaceous species")
```
Note that some curves are truncated. This is due to the limited values variation of the bioclimatic raster. 

We will now rescale the herbaceous climatic suitability * shading between 0 - 1, just for graphical purposes. 

```{r, eval=TRUE}
tmp <- shade.germ.r*my.second.species$suitab.raster
tmp <- rescale(tmp,  x.min = 0, x.max = 0.66, new.min = 0, new.max = 1)
my.second.species$suitab.raster <- tmp
```

Let's plot the two virtual species together
```{r, eval=TRUE}
par(mfrow=c(2,2))
plot(my.first.species$suitab.raster, main = "Virtual tree climatic suitability")
plot(my.second.species$suitab.raster, main = "Virtual herb climatic suitability")
plot(shade.germ.r, main = "Germination probability of the \n herbaceous virtual species")
plot(my.second.species$suitab.raster, main = "Virtual herb habitat suitability")
par(mfrow=c(1,1))
```

### Convert virtual species suitability indices to presence-absence
```{r, eval=TRUE, echo=FALSE,  message=FALSE}
new.pres.tree <- convertToPA(x = my.first.species,
                           # beta=0.55,
                           beta="random",
                           alpha = -0.05, plot = FALSE,
                           species.prevalence = 0.4)
```

Apply biotic and germination filter to herbaceous plant species climatic suitability
```{r, eval=TRUE, message=FALSE}
new.pres.herb<-convertToPA(x = my.second.species,
                            # beta=0.55,
                            beta = "random",
                            alpha = -0.05, plot = FALSE,
                            species.prevalence = 0.2)
```
```{r, eval=TRUE}
par(mfrow=c(1,2), oma=c(3,3,3,3))
plot(new.pres.tree$pa.raster, main = "Virtual tree species", legend = FALSE)
plot(new.pres.herb$pa.raster, main = "Virtual herbaceous species")
par(mfrow=c(1,1))
```

## 3. Structural equation modelling
```{r, eval=TRUE, message=FALSE}
library(dismo)
library(piecewiseSEM)
library(semEff)
```

### Training dataset preparation 
The complete survey of the two species is represented by the presence-absence raster files we generated using the *virtualspecies::convertToPA()* function. Since we rarely have a complete survey of the species presence-absence records in the geographic space, we displace randomly 1,000 points and retrieve information about the presence-absence of the two species along with the environmental conditions occurring in these locations.

```{r, eval=TRUE, echo=FALSE}
dat.sc<-readRDS("/home/ddare/working_files/ENM_MR/SEM/script/datasetSEM_20220401.RDS")
```
```{r, eval=FALSE}
{set.seed(123)
randpts <- sampleRandom(envData$bio1, size =  1000, sp = TRUE)}
crs(randpts) <- crs(new.pres.tree$pa.raster)
```
```{r}
# Stack variabels
pred.stack <- stack(envData$bio1, envData$bio12, 
                    shade.germ.r, new.pres.tree$pa.raster,
                    new.pres.herb$pa.raster)
names(pred.stack) <- c("bio1", "bio12", "herb.germ", "tree.pa",  "herb.pa"  )
```
```{r, eval=FALSE}
# Extract variables and scaling them 
dat <- na.omit(raster::extract(pred.stack, randpts, df = TRUE ))
dat.sc <- data.frame(scale(dat[,2:4], center = TRUE, scale = TRUE), dat[, 5:6])
```

### Assembling the SEM

Our causal framework implies that: 

* BIO1 and BIO12 have a direct effect on both the tree and herbaceous species occurrence (tree.pa, herb.pa), as well as on the herbaceous species germination (herb.germ);
* The herbaceous species germination (herb.germ) has a direct effect of the herbaceous species occurrence (herb.pa);
* The tree species occurrence (tree.pa) has a direct effect on the herbaceous species germination (herb.germ), and an indirect effect on the herbaceous species occurrence (herb.pa).

```{r}
vs.sem <- list(
  tree.pa = glm(tree.pa ~ bio1 + bio12 , data = dat.sc, family = binomial),
  herb.germ = glm(herb.germ ~ bio1 + bio12 + tree.pa ,  data = dat.sc),
  herb.pa = glm(herb.pa ~ bio1 + bio12 + herb.germ, data = dat.sc, family = binomial)
)

# Network visualization
plot(piecewiseSEM::as.psem(vs.sem, Class = "psem"))

# Note that you can also plot the causal diagram with a "tree" shape
# piecewiseSEM:::plot.psem(
#   piecewiseSEM::as.psem(vs.sem, Class = "psem"), #the values are the standardized estimates
#   node_attrs = data.frame(shape = "rectangle", color = "black", fillcolor = "grey"),
#   layout = "tree"
# )
```

#### Investigate SEM quality
Check for potential missing path(s) and update the SEM
```{r}
# Basis set = the smallest possible set of independence claims from a graph
basisSet(vs.sem) # Test if we are missing some connections and if those are significant 
dTable <- dSep(vs.sem) 
dTable
```

```{r}
fisherC(dTable)
```
We have a path that we did not consider in the SEM definition and that we would not include it because it is not-significant. Moreover, the Fisher's C test has a P>0.05, thus we did not reject H0 (the variance-covariance matrix of the model reflects the variance-covariance matrix of the data).

```{r, eval=TRUE}
psem_tmp <- piecewiseSEM::as.psem(vs.sem, Class = "psem")
mod.sum <- summary(psem_tmp, .progressBar = F, conserve = TRUE) 
kable(mod.sum$coefficients, align = "c")
kable(mod.sum$R2, align = "c")
```

### Calculate the effects
To calculate the effects, we first bootstrap and save the standardized direct effects. 
Provide a sufficient number of bootstraps samples, since a small number may produce NAs (see [this source](https://stats.stackexchange.com/questions/37918/why-is-the-error-estimated-adjustment-a-is-na-generated-from-r-boot-package)).
```{r, eval=TRUE, echo=FALSE}
vs.sem.boot<-readRDS("/home/ddare/working_files/ENM_MR/SEM/script/other_VS_vs.sem.boot_boot_2022-04-01.RDS")
```
```{r, eval=FALSE}
vs.sem.boot <- bootEff(psem_tmp, R = 1000,  seed = 13, parallel = "multicore", cl = 5)
```
```{r}
# Using the bootstrapped estimates we can now calculate direct, indirect, and total effects:
vs.sem.eff <- semEff(vs.sem.boot)
```
```{r}
# summary(vs.sem.eff, "herb.pa")
kable(vs.sem.eff$Summary$herb.pa, align = "c")
```

## 4. SEM Predictions
Now we can predict values for *herb.pa* using the *semEff::predEff()* function. 
As arguments, we need the model object and a RasterStack with the predictors in the region of interest. To save computational time, we will predict on a subset of the training area.

```{r}
# Crop rasters
pred.stack_red=crop(pred.stack[[c(1:4)]], extent(c(5, 17, 43, 49)) )

# Create predictions data.frame and scale the values
pred.dat <-as.data.frame(pred.stack_red, xy=TRUE)
pred.dat <- pred.dat[, c(3:6)]
pred.dat=data.frame(scale(pred.dat[,1:3]), "tree.pa"=pred.dat$tree.pa)
pred.dat=na.omit(pred.dat)
```

```{r, eval=FALSE}
# Extract the (total) effects of each variable to use for prediction
tot <- getTotEff(vs.sem.eff, "herb.pa")
tot.b <- getTotEff(vs.sem.eff, "herb.pa", type = "boot")

# SEM prediction
mod <- vs.sem$herb.pa
test_simple_pred <- predEff(mod, 
                            newdata = pred.dat, 
                            effects = tot, 
                            eff.boot = tot.b, 
                            type = "response")
```

```{r, eval=TRUE, echo=FALSE}
test_simple_pred<-readRDS("/home/ddare/working_files/ENM_MR/SEM/script/other_VS_simple_pred_2022-04-01.RDS")
```
```{r}
summary(test_simple_pred$fit)
```

Spatialize the predictions
```{r}
pred.dat <- as.data.frame(pred.stack_red, xy = TRUE)
pred.dat$id <- 1:nrow(pred.dat)
merge_id <- na.omit(pred.dat)$id

average_fit <- data.frame("id" = merge_id,
                       "average_fit" = test_simple_pred$fit,
                       "ci.lower" = test_simple_pred$ci.lower,
                       "ci.upper" = test_simple_pred$ci.upper )
pred.dat=pred.dat %>% 
  left_join(average_fit, by = "id")

r.fit <- data.frame(pred.dat[,1:2], "z" = pred.dat$average_fit)
r.fit <- raster::rasterFromXYZ(xyz = r.fit)
r.ci.low <- data.frame(pred.dat[,1:2], "z" = pred.dat$ci.lower)
r.ci.low= raster::rasterFromXYZ(xyz = r.ci.low)
r.ci.up <- data.frame(pred.dat[,1:2], "z" = pred.dat$ci.upper)
r.ci.up <- raster::rasterFromXYZ(xyz = r.ci.up)
```

Define color palette and legend range, then plot.
```{r}
# Set max and min values 
maxv <- 1
minv <- 0

# Set the breaks between min and max values
library(RColorBrewer)
brks <- seq(minv, maxv, by = 0.1)
nbrks <- length(brks)-1
r.range <- c(minv, maxv)

spect_pal <- colorRampPalette(rev(brewer.pal(11,'RdYlBu')))
ncols <- 10
spect_pal <- spect_pal(ncols)
```

```{r}
par(mfrow = c(2,2))
plot(crop(my.second.species$suitab.raster, extent(c(5, 17, 43, 49))  ), 
     main = "Virtual herbaceous plant \n (suitability + biotic filter)",
     breaks = brks,col=spect_pal,  
     legend = F, zlim=c(minv,maxv))
plot(r.ci.low, main="SEM CI low", 
     breaks = brks,
     col = spect_pal, 
     legend = F, 
     zlim = c(minv,maxv))
plot(r.fit, main="SEM prediction", 
     breaks = brks,
     col = spect_pal, 
     legend = F, 
     zlim = c(minv,maxv))
plot(r.ci.up, main = "SEM CI up", 
     breaks = brks,
     col = spect_pal, 
     legend = T, 
     zlim = c(minv,maxv))
par(mfrow = c(1,1))
```

## 5. Classic Generalized Linear Model (GLM)
We test also a classic SDM approach using a binomial GLM where the predictors are represented only by the bioclimatic variables and the tree species presence-absence is the response variable.

```{r}
myglm <- glm(herb.pa ~ tree.pa + bio1 + bio12, data = dat.sc, family = binomial)
summary(myglm)
print(paste("GLM Dsquared:", round(modEvA::Dsquared(myglm, adjust = TRUE),3)))
```
Let's plot the predictions
```{r}
# Predictions dataset
pred.dat.glm <-as.data.frame(pred.stack_red, xy = TRUE)
pred.dat.glm <- data.frame(scale(pred.dat.glm[, c(3:4, 6)]),
                           "tree.pa" = pred.dat.glm$tree.pa)
pred.dat.glm <- data.frame(na.omit(pred.dat.glm))

# Predict
glm.pred <- predict(myglm, newdata=pred.dat.glm, type="response")

# Spatialize the predictions
pred.dat.glm <- as.data.frame(pred.stack_red, xy = TRUE)
pred.dat.glm$id <- 1:nrow(pred.dat.glm)
merge_id <- na.omit(pred.dat.glm)$id
myglm_fit <- data.frame("id"=merge_id,"fit"=glm.pred)
pred.dat.glm <- pred.dat.glm %>% 
  left_join(myglm_fit, by="id")
glm.fit <- data.frame(pred.dat.glm[, 1:2], "z"=pred.dat.glm$fit)
glm.fit <- raster::rasterFromXYZ(xyz=glm.fit)

#plot
par(mfrow = c(1,2))
plot(crop(my.second.species$suitab.raster, extent(c(5, 17, 43, 49)) ),
     main = "herbaceous plant \n (suitability + biotic filter)",
     breaks = brks,
     col= spect_pal,
     legend = F, 
     zlim = c(minv,maxv))
plot(glm.fit, main = "Classic GLM prediction", 
     breaks = brks,
     col = spect_pal, 
     legend = T,
     zlim = c(minv,maxv))
par(mfrow = c(1,1))    
```
 
## 6. Comparison of SEM and GLM prediction accuracy metrics
We can qualitativerly compare the predictions of the two modelling techniques by plotting them.
```{r, message=FALSE}     
par(mfrow = c(2,2))
plot(crop(my.second.species$suitab.raster, extent(c(5, 17, 43, 49)) ),
  main = "herbaceous plant \n (suitability + biotic filter)",
  breaks = brks,
  col = spect_pal,
  legend = F, 
  zlim = c(minv,maxv))
plot(r.fit, main = "SEM prediction", 
     breaks = brks,
     col = spect_pal, 
     legend = F, 
     zlim = c(minv,maxv))
plot(glm.fit, main = "Classic GLM prediction", 
     breaks = brks,
     col = spect_pal, 
     legend = T, 
     zlim = c(minv,maxv))
plot(r.fit-glm.fit, main = "Delta SEM-GLM", legend = T, )
par(mfrow=c(1,1))
```

We also assess which modelling techniqu ( SEM or binomial GLM) exhibit the best predictive performance using the following metrics: Area Under the ROC Curve (AUC), Continuous Boyce Index, Sensitivity, Specificity, and True Skill Statistic (TSS). 
```{r, message=FALSE}
library(pROC)
library(ecospat)
out.stack <- stack(crop(new.pres.herb$pa.raster, extent(c(5, 17, 43, 49)) ), 
               r.fit, glm.fit )
names(out.stack) <- c("Observed", "SEM", "GLM")
```
```{r, message=FALSE, eval=TRUE, echo=FALSE}
# randpts <- sampleRandom(pred.stack_red$bio1, size =  500, sp = TRUE)
# crs(randpts) <- crs(new.pres.tree$pa.raster)
# saveRDS(randpts, "/home/ddare/working_files/ENM_MR/SEM/script/evalPoints_2022-04-29.RDS")
randpts<-readRDS("/home/ddare/working_files/ENM_MR/SEM/script/evalPoints_2022-04-29.RDS")
```
```{r, message=FALSE, eval=FALSE}
# Sample the geographic space
{set.seed(123)
randpts <- sampleRandom(out.stack$Observed, size =  500, sp = TRUE)}
crs(randpts) <- crs(new.pres.tree$pa.raster)
```
```{r, message=FALSE, eval=TRUE}
acc.metrics.df <- na.omit(raster::extract(out.stack, randpts, df = TRUE ))

pROC_obj_SEM <- roc(acc.metrics.df$Observed, acc.metrics.df$SEM,
                smoothed = TRUE,
                # arguments for ci
                ci = TRUE, ci.alpha = 0.9, stratified = FALSE,
                # arguments for plot
                plot = TRUE, auc.polygon = TRUE, max.auc.polygon = TRUE,
                grid = TRUE, print.auc = TRUE, show.thres = TRUE)

pROC_obj_GLM <- roc(acc.metrics.df$Observed, acc.metrics.df$GLM,
                smoothed = TRUE, ci = TRUE, ci.alpha = 0.9,
                stratified = FALSE, plot = TRUE, auc.polygon = TRUE,
                max.auc.polygon = TRUE,
                grid = TRUE, print.auc = TRUE, show.thres = TRUE)

```

Produce predictive accuracy metrics table
```{r, message=FALSE}
k <- rbind(data.frame("Model" = "SEM", 
                 "AUC" = round(auc(pROC_obj_SEM),3), 
                 round(coords(pROC_obj_SEM, "best",  transpose = FALSE, 
                              best.method="closest.topleft"),3), 
                 "CBI" = ecospat.boyce(fit = acc.metrics.df$SEM,
                                     obs = acc.metrics.df[which(acc.metrics.df$Observed==1),"SEM"],
                                     nclass = 0,
                                     window.w = "default", res = 100, PEplot = FALSE)$cor,
                 "TSS" = round(ecospat.max.tss(Pred = acc.metrics.df$SEM, 
                                             Sp.occ = acc.metrics.df$Observed)$max.TSS,3)),
      data.frame("Model" = "GLM", 
                 "AUC" = round(auc(pROC_obj_GLM),3), 
                 round(coords(pROC_obj_GLM, "best",  transpose = FALSE,
                              best.method="closest.topleft"),3), 
                 "CBI"= ecospat.boyce(fit = acc.metrics.df$GLM,
                                      obs = acc.metrics.df[which(acc.metrics.df$Observed==1),"GLM"],
                                      nclass = 0, window.w="default", res = 100, PEplot = FALSE)$cor,
                 "TSS"=round(ecospat.max.tss(Pred = acc.metrics.df$GLM,
                                             Sp.occ = acc.metrics.df$Observed)$max.TSS,3)))

kable(k, align = "c")

```

## Conclusions

Applying SEM to SDMs is conceptually and technically feasible (given that appropriate data are available).

The classic GLM approach got an R2 = `r round(modEvA::Dsquared(myglm, adjust = TRUE),2)`, while the SEM got an R2 = `r mod.sum$R2[which(mod.sum$R2$Response=="herb.pa"), "R.squared"]`, thus with a $\Delta$R2 of roughly `r mod.sum$R2[which(mod.sum$R2$Response=="herb.pa"), "R.squared"] - round(modEvA::Dsquared(myglm, adjust = TRUE),2)` in favor of the SEM model.  
